Estimating demographic bias on tests of children's early vocabulary

### 2022-01-16 paper/cogsci.Rmd
writing

While it is quite reasonable to expect that some CDI words are easier than others, and even to use these varying difficulties to predict variation in children's language ability, the use of psychometric models highlights the possibility that some CDI items may function differently (i.e., be more/less difficult) for different groups of children.
The idea that some items on a test may show bias, favoring one group over another, is known as Differential Item Function (DIF; @holland1993differential]. 
On any given test, it is clearly undesirable to have more items favoring one group (say, children from rural households) over another (urban children), as the test will overestimate the ability of test-takers in the former (rural) group -- and not because of any underlying mean difference in ability between the groups, but simply because the test is unfair.
A variety of statistical methods for detecting DIF have been proposed, and investigations have in several instances identified DIF for many items on tests favoring particular one group over another (e.g., rural vs. urban).
Our goal here is to test the items on the American CDI:WS for DIF along three main axes: sex (male vs. female), socioeconomic status (low- vs. high-SES), and race (white vs. non-white).

@@

# ethnicity
eth <- d_demo %>% filter(!is.na(eth_group), production!=0) %>%
  mutate(theta_eth = fscores(mod_intuitive_eth)) %>%
  filter(age<26, age>22) %>%
  rename(`Ethnic Group`=eth_group)
# cor.test(eth$production, eth$theta_eth) # .98

eth_m2 <- eth %>% group_by(`Ethnic Group`) %>%
  filter(age<26, age>22) %>%
  summarise(ability = mean(theta_eth, na.rm=T))
# now white=.52, nonwhite=.47 (was .59, .06)
p3 <- eth %>%
  ggplot(aes(x=theta_eth, color=`Ethnic Group`, fill=`Ethnic Group`)) +
    geom_density(alpha=.2) + xlab("Ability") + 
    geom_vline(data=eth_m, aes(xintercept=ability, color=`Ethnic Group`)) + 
    scale_color_discrete(type=eth_colors) + 
    scale_fill_discrete(type=eth_colors)

@@

define GLIMMER

>>>>

While it is quite reasonable to expect that some CDI words are easier than others, and even to use these varying difficulties to predict variation in children's language ability, the use of psychometric models highlights the possibility that some CDI items may function differently (i.e., be more/less difficult) for different groups of children.
The idea that some items on a test may show bias, favoring one group over another, is known as Differential Item Function (DIF; @holland1993differential]. 
On any given test, it is clearly undesirable to have more items favoring one group (say, children from rural households) over another (urban children), as the test will overestimate the ability of test-takers in the former (rural) group -- and not because of any underlying mean difference in ability between the groups, but simply because the test is unfair [@camilli2013ongoing].
A variety of statistical methods for detecting DIF have been proposed, and investigations have in several instances identified DIF for many items on tests favoring particular one group over another (e.g., rural vs. urban).
Our goal here is to test the items on the American CDI:WS for DIF along three main axes: sex (male vs. female), socioeconomic status (low- vs. high-SES), and race (white vs. non-white).

@@

# ethnicity
eth <- d_demo %>% filter(!is.na(eth_group), production!=0) %>%
  mutate(theta_eth = fscores(mod_intuitive_eth)) %>%
  filter(age<26, age>22) %>%
  mutate(Race=eth_group)
# cor.test(eth$production, eth$theta_eth) # .98

eth_m2 <- eth %>% group_by(Race) %>%
  filter(age<26, age>22) %>%
  summarise(ability = mean(theta_eth, na.rm=T))
# now white=.52, nonwhite=.47 (was .59, .06)
p3 <- eth %>%
  ggplot(aes(x=theta_eth, color=Race, fill=Race)) +
    geom_density(alpha=.2) + xlab("Ability") + 
    geom_vline(data=eth_m2, aes(xintercept=ability, color=Race)) + 
    scale_color_discrete(type=eth_colors) + 
    scale_fill_discrete(type=eth_colors)

@@

The first step we take to identify CDI items with DIF is to create GLIMMERs (Graphs of Logits Imputed Multiply with Means Equal; @stenhaug2021treading), which visualize between-item variation in group performance differences.

To give analysts a sense of the uncertainty about the existence of DIF, GLIMMER plots show distributions of parameter differences rather than point estimates. 
These distributions are generated by drawing 10,000 imputations from the item parameter covariance matrix.
If there is a cluster of items in the GLIMMER plot, the analyst may conclude that these items 
If there is no cluster, but rather a continuum of smoothly varying differences, the identification of DIF is fraught:
